---
title: "Grab bag of questions, answers, comments, and advice"
subtitle: "DFO TESA sdmTMB workshop"
author: ""
institute: ""
date: "November 4--6 2025"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->
```{r preamble, include=FALSE, cache=FALSE}
source(here::here("dfo-tesa-2025/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(sdmTMB)
```

# Can we use these models with fisheries dependent data?

- "Theoretically" no, in practice, yes. See "preferential sampling".

--

- Several papers show these can work very well for commercial CPUE and better than alternatives

--

- Theoretically, if all covariates driving the preferential sampling (e.g., depth, location) the sampling bias problem is ameliorated

--

- If worried, simulate it, or attempt to account for the preferential sampling.

---

# Things to be aware of when using environmental covariates

- Same principles apply with sdmTMB as any GLM/GLMM/GAM

--

- Ideally, use the same source for observations and grid, especially if or index standardization

--

- Covariates and random fields can compete to explain the same variance; can be seen a feature or a bug

--

- Don't be surprised if the fields explain a lot more variance than your fixed effects!

---

# Can sdmTMB be used with tracking data?

- Only sort of

--

- Not while taking advantage of all the information in the data

--

- If tracking data is reduced, yes

---

# Help me wrap my head around why we need the mesh and SPDE again

- Two things going on speed-wise: mesh and SPDE

--

- Mesh lets us work with (far) fewer random effects

--

- Mesh also gives us the matrices we need to do the SPDE math

--

- SPDE lets us work with the precision matrix instead of covariance matrix; avoids matrix inversion, much faster

--

- RTMB examples: [direct MVN Gaussian random field](https://github.com/seananderson/RTMB-TESA-spatial/blob/main/exercises/01-spatial-dmvnorm.R), [SPDE GMRF](https://github.com/seananderson/RTMB-TESA-spatial/blob/main/exercises/02-spatial-spde.R)

---

# What are common problems when starting out?

- Analysis paralysis with all the options

--

- Using (or starting with) too fine of a mesh

--

- Designing a mesh that doesn't cover your whole prediction area

--

- SDs that collapse to zero + poorly estimated range parameters

--

- Using different environmental data when fitting and predicting

---

# When should I use VAST, sdmTMB, or tinyVAST?

- `¯\_(ツ)_/¯`

--

- Timeline: SpatialDeltaGLMM, VAST, sdmTMB, tinyVAST

--

- Current actively developped packages: sdmTMB and tinyVAST

--

- If set up the same, VAST, sdmTMB, and tinyVAST can all fit *identical* models and have been extensively tested

---

## sdmTMB vs. tinyVAST (in my opinion!)

.small[
- Neither going anywhere soon; Jim and I work on both
]

--

.small[
- sdmTMB more user friendly, focusses on univariate responses
]

--

.small[
- sdmTMB has the most built-up set of functions for summarizing, checking, and working with the output
]

--

.small[
- sdmTMB can only fit "multivariate" responses if the fields are independent across categories
]
--

.small[
- tinyVAST focusses on multivariate responses but (of course) also does univariate responses
]

--

.small[
- Compared to VAST: adds structural equation and dynamic structural equation modelling
]

---

### Is there likely a need for High Performance Computing with sdmTMB?

- sdmTMB can do multicore likelihood calculations<br>`TMB::openmp(n=3, DLL="sdmTMB")`

--

- But it's not always faster!

--

- Memory can be an issue with larger datasets, and especially many (correlated) random effects (i.e., large meshes, AR1/random walks)

--

- For index standardization memory issues: <br>`?get_index_split`

--

- If fitting lots of models, absolutely!


